---
title: "4630 final project"
output: html_document
date: "2025-12-02"
---

```{r}
library(tidyverse)
library(rsample) # train/test split, CV
library(yardstick) # RMSE, R^2
library(glmnet) # ridge / lasso
library(tree) # regression tree
library(randomForest) # random forest

airbnb_raw <- read_csv("/Users/zhaoshibo/Desktop/STAT 4630/AB_NYC_2019.csv")
glimpse(airbnb_raw)

airbnb <- airbnb_raw %>%
  select(
    price,
    neighbourhood_group,
    neighbourhood,
    room_type,
    minimum_nights,
    number_of_reviews,
    reviews_per_month,
    calculated_host_listings_count,
    availability_365,
    latitude,
    longitude
  ) %>%
  # filters for price
  filter(price > 0, price <= 1000) %>%
  # log-transform price & set factor types
  mutate(
    log_price = log(price),
    neighbourhood_group = factor(neighbourhood_group),
    neighbourhood       = factor(neighbourhood),
    room_type           = factor(room_type),
    # set NA to 0
    reviews_per_month   = replace_na(reviews_per_month, 0)
  ) %>%
  drop_na()

glimpse(airbnb)
```

EDA
```{r}
# Distribution of price and log_price

airbnb %>%
ggplot(aes(price)) +
geom_histogram(bins = 50) +
labs(title = "Distribution of Airbnb Prices (capped at 1000)")

airbnb %>%
ggplot(aes(log_price)) +
geom_histogram(bins = 50) +
labs(title = "Distribution of log(price)")


# Boxplot by room type

airbnb %>%
ggplot(aes(room_type, price)) +
geom_boxplot() +
coord_cartesian(ylim = c(0, 500)) +  # avoid extreme tails
labs(title = "Price by Room Type (capped at 500)")

# Boxplot by borough

airbnb %>%
ggplot(aes(neighbourhood_group, price)) +
geom_boxplot() +
coord_cartesian(ylim = c(0, 500)) +
labs(title = "Price by Borough (neighbourhood_group)")

# Simple location scatter colored by borough

airbnb %>%
ggplot(aes(longitude, latitude, color = neighbourhood_group)) +
geom_point(alpha = 0.8, size = 0.8) +
labs(title = "Listing Locations by Borough") +
theme_minimal()

```

skewness so logged

Train / test split

```{r}
set.seed(4630)

#predict log_price

airbnb_model <- airbnb %>%
select(
log_price,
neighbourhood_group,
neighbourhood,
room_type,
minimum_nights,
number_of_reviews,
reviews_per_month,
calculated_host_listings_count,
availability_365,
latitude,
longitude
)

# stratify

split <- initial_split(airbnb_model, prop = 0.7, strata = neighbourhood_group)
train <- training(split)
test  <- testing(split)

nrow(train); nrow(test)

get_metrics <- function(truth, estimate) {
tibble(truth = truth,
estimate = as.numeric(estimate)) %>%
metrics(truth = truth, estimate = estimate) %>%
filter(.metric %in% c("rmse", "rsq"))
}

get_metrics <- function(truth, estimate) {
tibble(truth = truth,
estimate = as.numeric(estimate)) %>%
metrics(truth = truth, estimate = estimate) %>%
filter(.metric %in% c("rmse", "rsq"))
}

```

total sample size = 48645


MLR

```{r}
lm_fit <- lm(
log_price ~ neighbourhood_group + room_type +
minimum_nights + number_of_reviews + reviews_per_month +
calculated_host_listings_count + availability_365 +
latitude + longitude,
data = train
)

summary(lm_fit)


lm_pred <- predict(lm_fit, newdata = test)

lm_metrics <- get_metrics(test$log_price, lm_pred)
lm_metrics

```

The multiple linear regression model achieved an RMSE of 0.464 on log(price) and an R² of 0.519. This means the model explains about 52% of the variability in log-transformed Airbnb prices

An RMSE of 0.464 on the log scale corresponds roughly to a multiplicative factor of exp(0.464) = 1.59, meaning predictions are typically off by about 59% up or down in the original dollar prices.



Ridge & Lasso

```{r}
x_train <- model.matrix(
log_price ~ neighbourhood_group + neighbourhood + room_type +
minimum_nights + number_of_reviews + reviews_per_month +
calculated_host_listings_count + availability_365 +
latitude + longitude,
data = train
)[, -1]

x_test <- model.matrix(
log_price ~ neighbourhood_group + neighbourhood + room_type +
minimum_nights + number_of_reviews + reviews_per_month +
calculated_host_listings_count + availability_365 +
latitude + longitude,
data = test
)[, -1]

y_train <- train$log_price
y_test  <- test$log_price

dim(x_train); dim(x_test)

#Ridge
set.seed(123)
cv_ridge <- cv.glmnet(x_train, y_train, alpha = 0)
plot(cv_ridge)

bestlam_ridge <- cv_ridge$lambda.min
bestlam_ridge

ridge_pred <- predict(cv_ridge, s = bestlam_ridge, newx = x_test)
ridge_metrics <- get_metrics(y_test, ridge_pred)
ridge_metrics

ridge_coef <- coef(cv_ridge, s = bestlam_ridge)
ridge_coef[1:20, , drop=FALSE]

#Lasso
set.seed(123)
cv_lasso <- cv.glmnet(x_train, y_train, alpha = 1)
plot(cv_lasso)

bestlam_lasso <- cv_lasso$lambda.min
bestlam_lasso

lasso_pred <- predict(cv_lasso, s = bestlam_lasso, newx = x_test)
lasso_metrics <- get_metrics(y_test, lasso_pred)
lasso_metrics

lasso_coef <- coef(cv_lasso, s = bestlam_lasso)

# Show nonzero coefficients (excluding intercept)

lasso_coef[lasso_coef != 0]

```

Both Ridge and Lasso outperformed linear model and Lasso is slightly better than ridge


Regression Tree & Random Forest

```{r}
set.seed(123)
tree_fit <- tree(
  log_price ~ neighbourhood_group + room_type +
    minimum_nights + number_of_reviews + reviews_per_month +
    calculated_host_listings_count + availability_365 +
    latitude + longitude,
  data = train
)

summary(tree_fit)
plot(tree_fit); text(tree_fit, pretty = 0)

set.seed(123)
cv_tree <- cv.tree(tree_fit)
plot(cv_tree$size, cv_tree$dev, type = "b",
xlab = "Tree size", ylab = "CV deviance")

best_size <- cv_tree$size[which.min(cv_tree$dev)]
best_size

tree_pruned <- prune.tree(tree_fit, best = best_size)
plot(tree_pruned); text(tree_pruned, pretty = 0)

tree_pred <- predict(tree_pruned, newdata = test)
tree_metrics <- get_metrics(test$log_price, tree_pred)
tree_metrics

```

tree map: the random forest is splited by room type in the first stage then driven by borough and latitude


Random forest


